---
title: "Analyze Alaska deep sea coral and sponge data"
author: "Mike Sigler"
date: "3/25/2021"
output: html_document
---
##### steps completed prior to these analyses are: 
##### read and compile camera data, 
##### read and compile substrate data, 
##### match camera and substrate data, also adding rasters on variables (e.g., depth)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(PerformanceAnalytics)
library(ggfortify)
library(mgcv)
library(maps)
library(mapdata)
library(mapproj)
library(cowplot)
library(rgeos)
library(sf)
library(rnaturalearth)          #provides a map of countries of the entire world
library(rnaturalearthdata)
library(fpc)
library(cluster)
library(knitr)
```

### Read matched data (location, DSCS taxa, raster variables, e.g., depth, RockCobble)
```{r,echo=FALSE,cache=TRUE}
# Camera data
#  Location and depth of the camera deployment
#  DSCS taxa
#  substrate - RockCobble, proportion of rock or cobble habitat, from the camera data 
# Variables of interest (stored in rasters), average annual values
#  Bathy - depth
#  Bcurrent - current with tidal effect removed
#  Btemp - bottom temperature
#  Color - surface ocean color
#  Slope - slope
#  Tmax - maximum tidal current 

rm(list = ls())
# set project directories
projdir         <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\"
projdir.dat     <- "D:\\Documents\\AK Coral and Sponge Model\\Camera data\\"
projdir.results <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Results\\"

# read data
Location.DSCS <-select(as_tibble(read.csv(paste(projdir.results,"Location.DSCS.variables.csv",sep=""))),-X)
Species.List.DSCS <- select(as_tibble(read.csv(paste(projdir.results,"Species.List.DSCS.csv",sep=""))),-X)
variables <- c("Bathy","Bcurrent","Btemp","Color","Slope","Tmax","RockCobble") # update variable names
```

### Compile table of non-zero DSCS observations
```{r,echo=FALSE,cache=TRUE}
# Table 1 - count and mean of non-zero observations
Location.DSCS %>%                          # select DSCS data
  select(Acanthogorgia.sp:Porifera) ->    
  DSCS
  num.species <- nrow(Species.List.DSCS)
  num.records <- nrow(DSCS)
  Table.1 <- as.data.frame(cbind(Species.List.DSCS$Species.standard,rep("NA",num.species),rep("NA",num.species)))
  colnames(Table.1) <- c("Species.standard","Proportion.nonzero","Cpue.nonzero")

# compute count (proportion) and mean of non-zero observations
  for (spp in 1:num.species) {
    Table.1[spp,2] <- round(nrow(subset(DSCS,DSCS[,spp]>0))/num.records,3) # proportion
    temp <- subset(DSCS,DSCS[,spp]>0)
    Table.1[spp,3] <- round(sum(temp[,spp])/num.records,3)   # mean (consider computing median) 
  }
  Table.1 %>%
    arrange(desc(Proportion.nonzero)) ->  # order records in descending value
    Table.1
  kable(Table.1)
```

### Hierarchical cluster analysis
```{r,echo=FALSE,cache=TRUE}
# some input parameters (which are varied in the analyses) for sensitivity testing
num.groups <- 4                   # set number of groups
min.FO <- 0.05                    # set minimum frequency of occurrence

# in cluster analysis, examine only species above minimum FO
  Table.1 %>%                           
    filter(Proportion.nonzero>min.FO) ->
    Table.1

  Location.DSCS %>%                          
    select(Acanthogorgia.sp:Porifera) %>%    # select DSCS data
    select(Table.1[,1]) ->                   # select species above minimum FO from data set
    DSCS                                     # reduced data set (higher FO species)

# cluster analysis
#  https://www.statmethods.net/advstats/cluster.html
#  https://www.datacamp.com/community/tutorials/hierarchical-clustering-R
  
# Prepare Data
  DSCS <- na.omit(DSCS)                   # listwise deletion of missing records
  DSCS.scaled <- as_tibble(scale(DSCS))   # standardize variables to mean = 0 and sd = 1 
                                          # (but not normally distributed due to zeroes)

# Determine number of clusters
  wss <- (nrow(DSCS.scaled)-1)*sum(apply(DSCS.scaled,2,var))
  for (i in 2:15) wss[i] <- sum(kmeans(DSCS.scaled,
                                     centers=i)$withinss)
  plot(1:15, wss, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares") 
  
##############################################################################  
# Ward Hierarchical Clustering
  d <- dist(DSCS.scaled, method = "euclidean")   # distance matrix
  fitH <- hclust(d, method="ward.D")             # cluster analysis itself
  plot(fitH)                                     # display dendogram
  GroupH <- cutree(fitH, k=num.groups)           # cut tree into num.groups clusters
                                                 #   GroupH is the assigned group
  rect.hclust(fitH, k=num.groups, border="red")  # draw dendogram with red borders around the clusters 
  DSCS.cl <- mutate(DSCS.scaled, cluster = GroupH)  # append cluster assignment to scaled DSCS data
  count(DSCS.cl,cluster)                         # count cluster occurrences

# append cluster assignment to main data table 
  Location.DSCS.cl <- mutate(Location.DSCS, cluster = GroupH)  

# Create maps of the area of interest and plot the sampling locations
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
  theme_set(theme_light() + theme(legend.position = "top"))
  world <- ne_countries(scale = "medium", returnclass = "sf") 
  class(world)                                               
  
# New facet label names for clusters
  cluster.names <- c("Halipteris", "Low DSCS", "High DSCS", "Hexactinellid")
  Location.DSCS.cl$cluster <- factor(Location.DSCS.cl$cluster, levels = c("1", "2","3","4"),
                    labels = cluster.names)
# create maps
  ggplot(data = world) +
    geom_sf() +
    geom_point(
      data = Location.DSCS.cl,   #calls the location data
      aes(
        x= Start_long,
        y = Start_lat),
      size = 1, 
      shape = "circle", 
      colour = "red") +   
    coord_sf(
      xlim = c(-180,-140),      # western hemisphere    
      ylim = c(50,60.5), 
      expand = FALSE) ->
    p
  facet(p,facet.by="cluster") +
    theme(
      strip.text.x = element_text(
        size = 12, color = "black", face = "bold"))   # plot (facet) by cluster group :)

  ggplot(data = world) +
    geom_sf() +
    geom_point(
      data = Location.DSCS.cl,   #calls the location data
      aes(
        x= Start_long,
        y = Start_lat),
      size = 1, 
      shape = "circle", 
      colour = "red") +   
    coord_sf(
      xlim = c(160,180),        # eastern hemisphere
      ylim = c(50,60.5), 
      expand = FALSE) ->
    p
  facet(p,facet.by="cluster") +
    theme(
      strip.text.x = element_text(
        size = 12, color = "black", face = "bold"))          # plot (facet) by cluster group :)

# get cluster means
  meansH <- t(aggregate(DSCS.cl,by=list(DSCS.cl$cluster),FUN=mean))
  colnames(meansH) <- cluster.names   # name columns
  meansH <- meansH[-1,]               # drop first row (duplicates last row)
  write.csv(meansH,paste(projdir.results,"Hierarchical ",num.groups," groups ",min.FO, "min.FO",".csv",sep=""))
  kable(meansH)

# plot longitude and depth by cluster
  Location.DSCS.cl <- mutate(Location.DSCS.cl, Start_long360 = Start_long)  # initialize a column to store longitude 360 values
  for (j in 1:nrow(Location.DSCS.cl)) {
    if (Location.DSCS.cl$Start_long360[j]>0) 
      {Location.DSCS.cl$Start_long360[j] <- -180-(180-Location.DSCS.cl$Start_long360[j])}
  }

  ggplot(Location.DSCS.cl, aes(x=Start_long360, y = Depth, color = factor(cluster))) + 
    geom_point() +
    scale_y_reverse() ->
    p
  facet(p,facet.by="cluster")                                      # plot (facet) by cluster group :)

# Cluster Plot against 1st 2 principal components
# https://www.statmethods.net/advstats/cluster.html  

# vary parameters for most readable graph
  clusplot(DSCS.scaled, DSCS.cl$cluster, color=TRUE, shade=TRUE,
           labels=2, lines=0)

# Centroid Plot against 1st 2 discriminant functions
    plotcluster(DSCS.scaled, DSCS.cl$cluster) 
```

### compare measured depth (Depth) and extracted depth (Bathy)
```{r,echo=FALSE,cache=TRUE}
Location.DSCS %>%
  ggplot(aes(x=Depth,y=Bathy,color=region)) +
  geom_point() +
  geom_abline(slope=1,intercept=0)->
  p
p   # ==> some EBS discrepancies at ~150 m
    # Chris suspects a bad bathymetry area south of Pribilof Canyon
```

### boxplots of variables by cluster (scaled values)
```{r,echo=FALSE,cache=TRUE}
# choose columns
Location.DSCS.cl %>%
  select(Depth,Bcurrent:RockCobble) ->
  variables.cl
variables.cl <- scale(variables.cl)   # scale to Normal (0,1) (I think, need to check)
variables.cl <- cbind(Location.DSCS.cl[,2],variables.cl,Location.DSCS.cl[,43])

variables.cl %>%
  gather(variable,value,Depth:RockCobble) %>%
  ggplot(aes(x=cluster,y=value)) +
  geom_boxplot() +
  coord_flip() ->
  p
  facet(p,facet.by="variable")                                 
```
  
### compute median variable values by cluster
```{r,echo=FALSE,cache=TRUE}
  Location.DSCS.cl %>%
    select(all_of(variables)) %>%                 # select variables
    select(-c("Bathy")) ->                        # drop Bathy (instead use Depth, which was measured)
    temp1
  Location.DSCS.cl %>%
    select(c("cluster","Depth")) ->               # select cluster and Depth
    temp2
  var.clusters <- cbind(temp2,temp1)              # bind two tables together

  var.clusters %>%
    group_by(cluster) %>%
    summarise_all(list(median=median), na.rm = TRUE) ->   # compute variable medians by cluster
    var.medians
  kable(var.medians)
```
    
### Principal Components Analysis (PCA) of variables
```{r,echo=FALSE,cache=TRUE}
# first need to translate cluster group (alphabetical) to numeric
  cluster.col <- data.frame(cbind(cluster.names,seq(1,4)))
  colnames(cluster.col) <- c("cluster","cluster.number")
  var.clusters <- left_join(var.clusters,cluster.col,by = "cluster")
  
  var.clusters %>%
    filter(!is.na(RockCobble)) ->             # check for NA values 
    var.clusters
  nrow(var.clusters)                        # 862 records (2 RockCobble observations were NA)
  summary(var.clusters)                     # no NA values

# Correlation coefficients  
  chart.Correlation(select(var.clusters,-c("cluster","cluster.number")), histogram=TRUE, pch=19)  

# Principle components analysis
  PCA <- prcomp(select(var.clusters,-c("cluster","cluster.number")), scale=TRUE)     
  summary(PCA)                                                         

  autoplot(PCA, data = var.clusters, colour = 'cluster.number', shape = 1, 
           loadings = TRUE, loadings.colour = 'blue',
           loadings.label = TRUE, loadings.label.size = 3)            # color clusters, https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html
# Consider exploring   ggbiplot
``` 

### Values of RockCobble where each species occurs
```{r,echo=FALSE,cache=TRUE}
# Examine RockCobble distribution by species and color box plot by family
  Location.DSCS.cl %>%                                               # start with main data table
    gather(key=Species.standard,value=Density,Acanthogorgia.sp:Porifera) %>% # reformat data
    left_join(Species.List.DSCS,by="Species.standard") %>%                   # add family names
    filter (Density>0) %>%                                           # exclude zeros
    ggplot(aes(x=reorder(Species.standard,RockCobble,median),y=RockCobble,fill=Family)) +  
    geom_boxplot() +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)) +
    coord_flip() -> p1
  p1
  
# Examine RockCobble distribution by cluster
  Location.DSCS.cl %>%                                               # start with main data table
    ggplot(aes(x=reorder(cluster,RockCobble,median),y=RockCobble)) +  
    geom_boxplot() +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)) +
    coord_flip() -> p1
  p1
```

