---
title: "read and compile camera data"
author: "Mike Sigler"
date: "3/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(mgcv)
library(mapdata)
library(mapproj)
library(PerformanceAnalytics)  
```

### Read and compile camera data
```{r, echo=FALSE}
  rm(list = ls())

# survey.number - 1 through 7
#   survey.number            survey.name      number.transects
#              1 Data.1 Kodiak 2012-2014.csv               40
#              2        Data.2 MACE 2013.csv               64
#              3        Data.3 MACE 2015.csv               89
#              4        Data.4 MACE 2017.csv               74
#              5             Data.5 HAPC.csv               73
#              6        Data.6 Aleutians.csv              216
#              7         Data.7 EBS 2014.csv              250
#              8     AI SSL Density_data.csv               61
#
# Total number of transects - 867
# VARIABLES
# survey.number - added variable to identify data set (1 through 6)
# Deployment_ID - unique ID for each camera drop
# Start_long - DDD.DDDD
# End_long - DDD.DDDD
# Start_lat - DD.DDDDD
# End_lat - DD.DDDDD
# Depth - in meters
# distance_fished - in meters
# swath - in meters
# Group.2 and Species - lowest taxonomic level identified in original files, becomes "Species"
# Species.standard - lowest taxonomic level using Camera Data speciesList.csv (using standard naming)
# Group.3 - grouped taxonomic level in original file, deleted
# Species.group - grouped taxonomic level using Camera Data SpeciesList.csv
# x - count
# On.bottom - initial frame number (one per second) at start of counting; key field for time matching
# Off.bottom - final frame number (one per second) at start of counting; key field for time matching
# Speed - knots (I think)
# Percent_on (percent of time on bottom)
# Temperature (NO RECORDS in Gulf of Alaska data; records in Aleutian data)
# Species group names
#   coral
#   crab
#   fish
#   octopus
#   other
#   Pennatulacean
#   sponge
#   Stylasteridae
#   Zooplankton
# density (computed) (x/swath/distance) where swath and distance are meters

# set project directories
projdir <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\"
projdir.dat <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Camera data\\"
projdir.result <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\results\\"

###########################################################################
# Read data sets
# count the number of data files (camera surveys)
data.files.vector=list.files(projdir.dat)
nfiles <- length(data.files.vector)	            # 8 data files
sample_size <- data.frame(survey.number=integer(),   # create empty data frame, for storing sample size for each camera survey
                          survey.name=character(),
                          number.transects=integer())

# read each data file (camera survey) and store as e.g., Data.1
for (n in 1:nfiles)
{
  file.num <- data.files.vector[n]                                # nth data file
  file.name <- paste(projdir.dat,file.num,sep="")                 # name of nth data file
  Data <- as_tibble(read.csv(file.name))                          # Read nth data file
  sample_size[n,] <- c(n,file.num,nrow(count(Data,Deployment_ID)))  # Sample size of nth data file
  survey.number <- data.frame(survey.number=rep(n,nrow(Data)))    # Create data frame/column for survey number
  Data <- bind_cols(survey.number,Data)                           # Add survey number to Data
  assign(paste("Data.",n,sep=""),Data)                            # Assign individual data frame names for each data set
}

  sample_size                        # Sample size by camera survey
  sum(as.numeric(sample_size[,3]))   # Total sample size

#####################################################################  
# Combine surveys into one table, mydata
#  start with the 3 MACE surveys (Data.2, Data.3, Data.4)
  str(Data.2)
  str(Data.3)   # Data.2 and Data.3 have the same format
  mydata <- bind_rows(Data.2,Data.3)  # combine data sets 2 and 3
  mydata$distance_fished <- mydata$Percent_on *
                            mydata$distance_fished   # Both Data.2 and Data.3 have values for Percent_on. 
                                                     # Chris said to adjust distance_fished for Percent_on
  mydata$Percent_on <- NA                            # now change Percent_on to NA, since adjustment completed
                # Data.4
  str(Data.4)   # compared to Data.2/3
                #   drops speed column and adds temperature column (but no recorded temperatures) 
                #     ==> add Temperature column to Data.2/3 and Speed column to Data.4 (with NA values)
                #   EventID (a number) substituted for Haul (same value as Deployment_ID)  
                #     ==> drop Event_ID from Data.4 and Haul from Data.2/3
                #   no Percent.on   
                #     ==> add Percent_on with NA values to Data.4 
                #   adds a column Swath  
                #   from Chris, 10/21/2020: I looked at the Swath column in my code. It turns out for 2017 you should use 
                #     the “Swath” column. It should only appear in that year. I had to go back and do some manual 
                #     calculations on that one, which is the reason for the multiple columns.
                #     ==> rename column Swath to swath (first deleting original swath) for Data.4
                #   adds a column for species group (Group.2) to specific species (Group.3)
                #     but Data.2/3 only has a column for specific species (Group.2, SO DIFFERENT NAMING)
                #     ==> switch column names in Data.4 to match Data.2/3 (Group.2 = specific species) 
                #     ==> add column for species group (Group.3) to Data.2/3 (filled with NA, ***FOR NOW***)
                # Chris says that Isididae in fact is a misidentified Primnoidae
                # changed in the input file, Deployment_ID=="D20170622-T090107"
                # "Isididae" ==> "Primnoidae"
  
# data revisions to Data.4 to match Data.2 and Data.3  
  mydata$Temperature <- NA
  Data.4$Speed <- NA
  Data.4 <- select(Data.4,-Event_ID)
  mydata <- select(mydata,-Haul)
  Data.4$Percent_on <- NA
  
  Data.4 <- select (Data.4,-swath)   # drop swath (the original value) from Data.4
  Data.4$swath <- Data.4$Swath       # assign adjusted value (Swath) in place of the original value
  Data.4 <- select(Data.4,-Swath)    # drop Swath (now duplicates swath)

  Data.4$temp <- Data.4$Group.2      # place species group data in a temporary column
  Data.4$Group.2 <- Data.4$Group.3   # rename species specific data as Group.2
  Data.4$Group.3 <- Data.4$temp      # rename temp (species group) data as Group.3
  Data.4 <- select(Data.4,-temp)     # remove temp column
  mydata$Group.3 <- NA  
# check data revisions
  str(mydata); str(Data.4)           # looks good (matching names and 18 columns each)
  
# add data set #4 to mydata (data sets #2 and #3)
  mydata <- bind_rows(mydata,Data.4)
  dim(mydata)                        # looks good
  summary(mydata)                    # no temperatures recorded in the MACE surveys!
  
# Add Kodiak data (Data.1)
  str(Data.1);str(mydata)            # all data columns in Data.1 are found in mydata with the same variable naming 
  dim(Data.1);dim(mydata)            # fewer columns (11) in Data.1 than mydata (18)
  summary(Data.1)                    # again, no temperatures recorded :(
  mydata <- full_join(Data.1,mydata) # missing columns in Data.1 added with NA values
  dim(mydata)                        # looks good

# Add HAPC data (Data.5)
  str(Data.5);str(mydata)  
  summary(Data.5)
# compared to mydata
  #   uses distance instead of distance_fished 
  #     ==> change name
  #   uses count instead of x 
  #     ==> change name
  #   uses density with no equivalent in mydata 
  #     ==> drop density (will be computed later, compare these values to computation)
  #   missing columns: On.bottom, Off.bottom, Speed, Percent_on, Temperature, Group.3
  #     ==> add columns
# data revisions    
  Data.5$distance_fished <- Data.5$distance   
  Data.5$x <- Data.5$count              
  Data.5 <- select(Data.5,-distance)
  Data.5 <- select(Data.5,-count)
  Data.5 <- select(Data.5,-density)   ### CHECK COMPUTATION CAREFULLY ###
  Data.5$On.bottom <- NA
  Data.5$Off.bottom <- NA
  Data.5$Speed <- NA
  Data.5$Percent_on <- NA
  Data.5$Temperature <- NA
  Data.5$Group.3 <- NA  

# check revisions
  str(Data.5);str(mydata)  

# combine data sets
  test <- bind_rows(mydata,Data.5)
  dim(test);dim(mydata);dim(Data.5)   # looks good
  mydata <- bind_rows(mydata,Data.5)  # mydata includes all Gulf of Alaska data sets (#1-5)
  
# Add the Aleutians data (Data.6)
  str(Data.6);str(mydata)  
  summary(Data.6)
# compared to mydata
#   has additional column 'Haul' (as well as 'Deployment_ID')
#     ==> drop haul
#   distance_fished has not been adjusted for percent_on in Data.6, but has been adjusted in mydata
#     ==> adjust distance_fished for percent_on
#   Group.3 is the lowest taxonomic level identified in Data.6,
#     whereas Group.2 is the lowest taxonomic level in mydata
#     In Data.6, drop Group.2 and rename Group.3 to Group.2
#   No speed column 
#     ==> add speed column to Data.6
#   Count is used in Data.6 and x in mydata
#     ==> rename Count to x in Data.6
# data revisions    
  Data.6 <- select(Data.6,-Haul)     # drop haul
  Data.6$distance_fished <- Data.6$distance_fished * Data.6$Percent_on # adjust distance fished by percent on bottom
                                     # rename groups
  Data.6$temp <- Data.6$Group.2      # place species group data in a temporary column
  Data.6$Group.2 <- Data.6$Group.3   # rename species specific data as Group.2
  Data.6$Group.3 <- Data.6$temp      # rename temp (species group) data as Group.3
  Data.6 <- select(Data.6,-temp)     # remove temp column
  Data.6$Speed <- NA
  Data.6$x <- Data.6$Count           # rename count            
  Data.6 <- select(Data.6,-Count)
  Data.6$test <- Data.6$x / Data.6$distance_fished / Data.6$swath 
                                     # check density computation: match density values in the original file
  Data.6 <- select(Data.6,-Density)  # remove original density value
  Data.6 <- select(Data.6,-test)     # remove test value
    
  # combine data sets
  test <- bind_rows(mydata,Data.6)
  dim(test);dim(mydata);dim(Data.6)   # looks good
  mydata <- bind_rows(mydata,Data.6)  # mydata includes all Gulf of Alaska data sets (#1-5) 
                                      #   and the Aleutian Islands data (#6)
  
# Add the eastern Bering Sea data (Data.7)
  str(Data.7);str(mydata)  
  summary(Data.7)
  # compared to mydata, Data.7 ...
  #   has additional column 'Haul' (as well as 'Deployment_ID')
  #     ==> drop haul
  #   distance_fished has been adjusted for percent_on in Data.7 (computed density and compared to Data.7 values)
  #     ==> no need to adjust distance_fished for percent_on
  #   missing 'Speed'
  #     ==> add Speed
  #   Group.2 is the lowest taxonomic level identified in Data.7, same as mydata
  #     Group.3 is identified in Data.7 but not in mydata
  #     ==> add Group.3 (with NA values) to Data.7
  #   Area and Density are included in Data.7 but not my data
  #     ==> drop these columns from Data.7
# data revisions    
  Data.7 <- select(Data.7,-Haul)     # drop haul
  Data.7$Group.3 <- NA               # add Group.3
  Data.7$Speed <- NA                 # add Speed
  Data.7 <- select(Data.7,-Area)     # drop Area
  Data.7 <- select(Data.7,-Density)  # drop Density
  
# combine data sets
  test <- bind_rows(mydata,Data.7)
  dim(test);dim(mydata);dim(Data.7)   # looks good
  mydata <- bind_rows(mydata,Data.7)  # mydata includes all Gulf of Alaska data sets (#1-5), 
                                      #   the Aleutian Islands data (#6), and 
                                      #   the eastern Bering Sea data (#7)
  
# Add the Aleutians data collected within Steller sea lion protection zones (Data.8)
  str(Data.8);str(mydata)  
  summary(Data.8)
  
# Data.8 has two extra columns, taxa and density and is missing several columns (e.g., On.bottom)
# data revisions    
  Data.8 <- select(Data.8,-c(Taxa,Density))     # drop haul
  Data.8$On.bottom <- NA             # add On.bottom
  Data.8$Off.bottom <- NA            # add On.bottom
  Data.8$Speed <- NA                 # add Speed
  Data.8$Percent_on <- NA            
  Data.8$Temperature <- NA           
  Data.8$Group.3 <- NA               # add Group.3

# combine data sets
  test <- bind_rows(mydata,Data.8)
  dim(test);dim(mydata);dim(Data.8)   # looks good
  mydata <- bind_rows(mydata,Data.8)  # mydata includes all Gulf of Alaska data sets (#1-5), 
                                      #   the Aleutian Islands data (#6), and 
                                      #   the eastern Bering Sea data (#7), and
                                      #   Aleutians sea lion protection areas
 nrow(mydata)                         # 7,960 records
 
 # check whether location is present
 summary(mydata)                      # 24 records with location and depth missing
 mydata %>%                           # delete those records
   filter(!is.na(Start_long)) ->
   temp
 nrow(temp)
 mydata <- temp
 nrow(mydata)
###################################################################################################
# Check data
# Check for rows with NA for x (the count column), i.e., camera deployed but no organisms observed
  filter(mydata,is.na(x)) # 9 rows, for all cases, only one record with no Group.2 observations (NA)
#  survey.number     Deployment_ID 
#  2 D20130706-T095847   
#  2 D20130731-T100634
#  2 D20130731-T105057
#  3 D20150719-T083545 
#  6 D20140506-T214001   
#  6           Haul 13 
#  7 D20140818-T163125   
#  8 SSL_2016_HAUL1 
#  8    SSL_2016_HAUL5
  ### CHRIS: treat as zeros ###
#  ==> for these 9 cases, rename Group.2 to "None" and set x = 0
  mydata$x[is.na(mydata$Group.2)] <- 0
  mydata$Group.2[is.na(mydata$Group.2)] <- "None"
  nrow(mydata)
  filter(mydata,is.na(x)) # 0 rows
  filter(mydata,Group.2 == "None")   # 0 rows, correct
    
# reduce and rename columns 
  mydata <- select(mydata,-On.bottom,-Off.bottom,-Speed,-Percent_on)   # columns not used in analyses
  mydata$Species <- mydata$Group.2             # rename Group.2 (original species name)
  mydata <- select(mydata,-c(Group.2,Group.3)) # remove Group.2 and Group.3, 
                                 #   lists of species names which will be standardized in the next section
  colnames(mydata)
```

### Check species names, add new names to the master list, SpeciesList.revised.V3.csv
#### RERUN ONLY IF new data added
```{r, echo=FALSE}
#   run the next several steps only if new data added (0/1 toggle)
# At Chris's advice, standard name for "Halipteris willemoesi" and "Pennatulacean" is "Halipteris sp"
  run.sppcheck <- 0
  if (run.sppcheck == 1) {
  mydata %>%
    group_by(Species) %>%
    summarise(avg = mean(x)) %>%
    arrange(avg) ->        
    summary.mydata.Species                                            # list of Species
    summary.mydata.Species <- select(summary.mydata.Species,-avg)     # select only the list of Species
  Species.List <- as_tibble(read.csv(paste(projdir.result,"Species.List.csv",sep="")))   # load table of standard names

  temp <- left_join(summary.mydata.Species,Species.List,by="Species") # compare lists (missing names show as "NA") 
  temp %>%
    filter(is.na(Species.standard))
                                                                      # ==> add missing names to master list
  
  nrow(Species.List)                                               # 491 records with some duplicates
  Species.List <- distinct(Species.List)                           # remove matching rows in Species.List
  nrow(Species.List)                                               # 491 records

# bookkeeping step: reduce Species.List to only names found in mydata
  temp <- left_join(mydata,Species.List,by="Species")              # match standard names to mydata,
  nrow(temp); nrow(mydata)                                         # distinct command worked!
  mydata <- temp                                                   # rename

  mydata %>%
    select(c("Species.standard","Family","Species.group")) %>%
    distinct() ->
    Species.List
  nrow(Species.List)   #  82 Standard.species in mydata, Gulf of Alaska
                       # 106 Standard.species in mydata, adding Aleutian Islands
                       # 134 Standard.species in mydata, adding eastern Bering Sea
                       # 138 Standard.species in mydata, adding AI sea lion protection areas
  } 
```

### Standardize species, family and group names in mydata
```{r, echo=FALSE}
  Species.List <- read.csv(paste(projdir.result,"Species.List.csv",sep=""))    # Load final version of Species.List  
  mydata <- left_join(mydata,Species.List,by="Species")     # Add standard names
  mydata <- select(mydata,-Species)                         # Drop non-standard species name
  mydata %>%                                                # reduce species list to those found in mydata
    select(c("Species.standard","Family","Species.group")) %>%
    distinct() ->
    Species.List
  nrow(Species.List)   # 138 Standard.species in mydata 
  
# reorder columns
  colnames(mydata)
  col.order <- c("survey.number","Deployment_ID","Start_long","Start_lat","End_long","End_lat","Depth","Temperature",
                 "distance_fished","swath","Species.standard","Family","Species.group","x") 
  mydata <- mydata[, col.order]
  colnames(mydata)   # looks good
  nrow(mydata)       # 2163 records for Gulf of Alaska,
                     #   increased to 4404 records with addition of the Aleutians data,
                     #   increased to 6861 records with addition of the eastern Bering Sea data  
                     #   increased to 7936 records with addition of AI SSL data

# with addition of the Aleutians data, Deployment_ID no longer is unique
#   haul number used in some records for Data.1 and all records for Data.6,
#   some haul numbers in Data.6 duplicate those in Data.1
#   add data set number at start of Deployment_ID
  nrecords <- nrow(mydata)
  for (n in 1:nrecords)  {
    mydata$Deployment_ID[n] <- paste("Data.",mydata$survey.number[n],".",mydata$Deployment_ID[n],sep="")
  }
  nrow(mydata)       # 7936 records   
  
# add a columnn for region (e.g., GOA) to mydata
  regions <- bind_cols(seq(1,8,1),
            c(rep("GOA",5),"AI","EBS","AI") )
  colnames(regions) <- c("survey.number","region")
  mydata <- left_join(mydata,regions,by="survey.number")
  nrow(mydata)       # 7936 records     
### ==> final version of mydata ###
```

### Tabulate the data by camera deployment and grouped by species, family and species group (e.g., fish) 
```{r, echo=FALSE}
# tidy the data (each data set in a tibble, each variable in a column)
# by standard species name
  mydata %>% 
    group_by(Deployment_ID,Species.standard) %>%
    summarise(sum=sum(x)) %>%
    spread(Species.standard,sum,fill=0) ->
    Count.Speciesstandard
  summary(Count.Speciesstandard)
   
# by family
  mydata %>% 
    group_by(Deployment_ID,Family) %>%
    summarise(sum=sum(x)) %>%
    spread(Family,sum,fill=0) ->
    Count.Family
  summary(Count.Family)
  
# by species group
  mydata %>% 
    group_by(Deployment_ID,Species.group) %>%
    summarise(sum=sum(x)) %>%
    spread(Species.group,sum,fill=0) ->
    Count.Speciesgroup  
  summary(Count.Speciesgroup)
```
   
### Group data by camera deployment and compute density for all species (coral, sponge, fish, etc.)  
```{r, echo=FALSE}
# location information
  mydata %>%
    select(c(survey.number,Deployment_ID,Start_long,Start_lat,End_long,End_lat,Depth,distance_fished,swath,region)) %>%
    distinct() ->
    Location
  Location$area_size <- Location$distance_fished * Location$swath
  nrow(Location)          # 866 camera drops

# check Location information
#   on-bottom typically 15 minutes except HAPC dives which were typically 5 minutes and as short as 2 minutes (when seafloor sandy)  
#   distance fished ~250-300 m; longer distances usually are high current areas (faster drift)
  summary(Location$distance_fished)
  summary(Location$swath)

# plot distance_fished by survey
  Location %>%
    ggplot(aes(x=distance_fished)) +
    geom_histogram(bins = 20) +
    ggtitle("Survey 1") ->
    p
  facet(p,facet.by="survey.number")                                     
  
# compute density (per hectare = 10000 m^2)
#   by standard species
  Density.Speciesstandard <- Count.Speciesstandard                          # Initialize density table
  Count.Speciesstandard <- left_join(Count.Speciesstandard,Location,by="Deployment_ID")   # match sampled areas to count data
  Density.Speciesstandard[,2:ncol(Density.Speciesstandard)] <-
    Count.Speciesstandard[,2:ncol(Density.Speciesstandard)]/                # select columns with species counts
    Count.Speciesstandard$area_size *                                            # divide by area (m^2)
    10000                                                                   # per 10000 m^2 = 1 hectare  

#   by family
  Density.Family <- Count.Family                                            # Initialize density table
  Count.Family <- left_join(Count.Family,Location,by="Deployment_ID")       # match sampled areas to count data
  Density.Family[,2:ncol(Density.Family)] <-
    Count.Family[,2:ncol(Density.Family)]/                                  # select columns with species counts
    Count.Family$area_size *                                                     # divide by area (m^2)
    10000                                                                   # per 10000 m^2 = 1 hectare  

# by species groups
  Density.Speciesgroup <- Count.Speciesgroup                                # Initialize density table
  Count.Speciesgroup <- left_join(Count.Speciesgroup,Location,by="Deployment_ID")   # match sampled locations to count data
  Density.Speciesgroup[,2:ncol(Density.Speciesgroup)] <-
    Count.Speciesgroup[,2:ncol(Density.Speciesgroup)]/                      # select columns with species counts
    Count.Speciesgroup$area_size *                                               # divide by area (m^2)
    10000                                                                   # per hectare  
  
# verify density computation by comparison with Data.4 (where density was already computed in the file)
#  Density.Speciesstandard$'Flatfish unid'[Density.Speciesstandard$Deployment_ID=="HAPC_19"]     
                                                                            # ==> matches 0.049751941 m-2 in HAPC_data.csv
#  ==> 7 tables
# count:        Count.Speciesstandard,  Count.Family, Count.Speciesgroup
# density:      Density.Speciesstandard, Density.Family, Density.Speciesgroup
# location:     Location
```

### Summarize deep sea coral and sponge data, 866 camera drops, and write to results
```{r, echo=FALSE}
#   first make list of coral, sponge, sea whip/pen, hydrozoan (Stylasteridae) standard species found in mydata
  Species.List %>%
    filter(Species.group=="Sponge"|Species.group=="Coral"|
             Species.group=="SeaWhipPen"|Species.group=="Hydrozoan") %>%
    arrange(Species.group,Family,Species.standard) ->           
    Species.List.DSCS
  nrow(Species.List.DSCS)   # 18 standard species in the Gulf of Alaska, 
                            #   24 by adding the Aleutian Islands data, 
                            #   26 by adding the eastern Bering Sea data,
                            #   27 by adding AI SSL data
  write.csv(Species.List.DSCS,paste(projdir.result,"Species.List.DSCS.csv",sep=""))   # list of coral, sponge, sea whip/pen Species names in new camera data

# then select columns with coral, sponge or sea whip/pen
#   there may be a more efficient way to do this, e.g., list the colnames and then select based on the colname list
  spp <- 1
  Density.Speciesstandard %>%
    select(Species.List.DSCS$Species.standard[spp]) ->
    DSCS
  for (spp in 2:nrow(Species.List.DSCS)) {
  Density.Speciesstandard %>%
    select(Species.List.DSCS$Species.standard[spp]) ->
    temp
  DSCS <- cbind(DSCS,temp[,2])
  }
  summary(DSCS)    
  
# table of coral, sponge, sea whip/pen      
# Table 1 - count and mean of non-zero observations
  num.species <- nrow(Species.List.DSCS)
  num.records <- nrow(DSCS)
  Table.1 <- as.data.frame(cbind(Species.List.DSCS$Species.standard,rep("NA",num.species),rep("NA",num.species)))
  colnames(Table.1) <- c("Species.standard","Proportion.nonzero","Cpue.nonzero")
# compute count (proportion) and mean of non-zero observations
  # use loops again
  for (spp in 1:num.species) {
    Table.1[spp,2] <- round(nrow(subset(DSCS,DSCS[,spp+1]>0))/num.records,3)
    temp <- subset(DSCS,DSCS[,spp+1]>0)
    Table.1[spp,3] <- round(sum(temp[,spp+1])/num.records,3)   # consider computing median 
  }
  Table.1 %>%
    arrange(desc(Proportion.nonzero)) ->
    Table.1
#  write.csv(Table.1,paste(projdir.result,"Table.1.csv",sep=""))
  
# Output file for higher level analyses  
  Location.DSCS <- left_join(Location,DSCS,by="Deployment_ID")         # join location and cpue data
  nrow(Location.DSCS)  
  write.csv(select(Location.DSCS,-c(distance_fished,swath,area_size)),
            paste(projdir.result,"Location.DSCS.csv",sep=""))                                       # write to file, excluding sample effort info
```

### Some data plots
#### some (but not all) are duplicated in analysis.Rmd
```{r, echo=FALSE}  
# Bar plots (bp) by species
  bp.proportion <- ggbarplot(Table.1, x = "Species.standard", y = "Proportion.nonzero",
                    color = "black",            # Set bar border colors to white
                    palette = "jco",            # jco journal color palett. see ?ggpar
                    sort.val = "asc",           # Sort the value in ascending order
                    x.text.angle = 90           # Rotate vertically x axis texts
    )
  bp.proportion

  Table.1$Cpue.nonzero <- as.numeric(Table.1$Cpue.nonzero)  
  bp.cpue <- ggbarplot(Table.1, x = "Species.standard", y = "Cpue.nonzero",
                       color = "black",            # Set bar border colors to white
                       palette = "jco",            # jco journal color palett. see ?ggpar
                       sort.val = "asc",           # Sort the value in ascending order
                       x.text.angle = 90           # Rotate vertically x axis texts
  )
  bp.cpue

########################################################################
# Examine sampling intensity by depth and longitude
#   first need longitude on a single scale 
  Location$Start_long360 <- Location$Start_long  # initialize a column to store single scale (0-360) values
  for (j in 1:nrow(Location)) {
    if (Location$Start_long360[j]>0) {Location$Start_long360[j] <- -180-(180-Location$Start_long360[j])}
  }
  summary(Location$Start_long360)   # computation works
  Location %>%
    ggplot(aes(x=Depth)) +        # by depth
    geom_histogram() ->
    p
  facet(p,facet.by="region")
  
  Location %>%
    ggplot(aes(x=Start_long360)) +   # by longitude
    geom_histogram() ->
    p
  facet(p,facet.by="region")

  Location %>%
    ggplot(aes(x=Start_lat)) +   # by latitude
    geom_histogram() ->
    p
  facet(p,facet.by="region")
####################################################################    
# Examine depth distribution by species 
#   Scatterplot
  Location %>%                                                     # start with Location table
    select(Deployment_ID,Depth) %>%                                # select deployment_ID and depth
    left_join(DSCS,by="Deployment_ID") %>%                         # join density data
    gather(key=Species,value=Density,Acanthogorgia.sp:Porifera) %>%   # reformat data
    filter (Density>0) %>%                                         # include only non-zero densities
    ggplot(aes(x=Depth,y=Density)) +                               # plot density vs. depth
    geom_point(shape=1) +                                          # scatterplot
    geom_smooth() ->                                               # add loess smoother
    p
  facet(p,facet.by="Species")                                      # plot (facet) by species :)

# boxplot 
  Location %>%                                                               # start with Location table
    select(Deployment_ID,Depth) %>%                                          # select deployment_ID and depth
    left_join(DSCS,by="Deployment_ID") %>%                                   # join density data
    gather(key=Species.standard,value=Density,Acanthogorgia.sp:Porifera) %>%    # reformat data
    left_join(Species.List.DSCS,by="Species.standard") %>%                   # add family names
    filter (Density>0) %>%                                                   # include only non-zero densities
    ggplot(aes(x=reorder(Species.standard,Depth,median),y=Depth,fill=Family)) +    # plot depth by species and color by family
    geom_boxplot() +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)) +
    coord_flip() -> p1
  p1
  
##############################################################################
  # Examine longitude distribution by species 
  #   Scatterplot
  Location %>%                                                     # start with Location table
    select(Deployment_ID,Start_long360) %>%                           # select deployment_ID and start longitude
    left_join(DSCS,by="Deployment_ID") %>%                         # join density data
    gather(key=Species,value=Density,Acanthogorgia.sp:Porifera) %>%   # reformat data
    filter (Density>0) %>%                                         # include only non-zero densities
    ggplot(aes(x=Start_long360,y=Density)) +                               # plot density vs. depth
    geom_point(shape=1) +                                          # scatterplot
    geom_smooth() ->                                               # add loess smoother
    p
  facet(p,facet.by="Species")                                      # plot (facet) by species :)
  
# boxplot 
  Location %>%                                                               # start with Location table
    select(Deployment_ID,Start_long360) %>%                                          # select deployment_ID and depth
    left_join(DSCS,by="Deployment_ID") %>%                                   # join density data
    gather(key=Species.standard,value=Density,Acanthogorgia.sp:Porifera) %>%    # reformat data
    left_join(Species.List.DSCS,by="Species.standard") %>%                   # add family names
    filter (Density>0) %>%                                                   # include only non-zero densities
    ggplot(aes(x=reorder(Species.standard,Start_long360,median),y=Start_long360,fill=Family)) +         # plot depth by species and color by family
    geom_boxplot() +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)) +
    coord_flip() -> p2
  p2
```
