---
title: "read trawl data and analyze"
author: "Mike Sigler"
date: "4/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### read and analyze Gulf of Alaska trawl survey data for deep sea coral and sponge

### read libraries
```{r,cache=TRUE,echo=FALSE}

## load libraries
library(mgcv)
library(PresenceAbsence)
library(gbm)
library(dismo)
library(raster)
library(rgdal)
library(maptools)
library(gstat)
library(rgeos)
library(proj4)
library(sp)
library(maptools)
library(maps)
library(xtable)
library(randomForest)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(PerformanceAnalytics)
library(ggfortify)
library(lubridate)
```

### set up directories
```{r, cache=TRUE, echo=FALSE}
#rm(list = ls())
# set project directories
data.path     <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Trawl data\\"
#GOArasters.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Variables_GOA_1km\\"
#AIrasters.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Variables_AI_1km_1000m\\"
results.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Results\\"
predictionrasters.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Results\\PredictionRasters\\"
projectionrasters.GOA.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\ProjectionRastersGOA\\"
projdir.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\"
```

### read trawl data
```{r, cache=TRUE, echo=FALSE}
# read trawl data
GOA_data <-as_tibble(read.csv(paste(data.path,"GOA_Trawl_data.csv",sep="")))
GOA_data$year <- year(mdy_hms(GOA_data$START_TIME))

# some data checks
# cross tabulate
GOA_data %>%
  count(Taxon)   # looks good
# locations
GOA_data %>%
  ggplot(aes(x=START_LONGITUDE)) + geom_histogram()  # looks good
GOA_data %>%
  ggplot(aes(x=START_LATITUDE)) + geom_histogram()   # looks good
GOA_data %>%
  count(year)   # looks good

# Group data by haul
# Taxon CPUE
  GOA_data %>% 
    group_by(HAULJOIN,Taxon) %>%
    summarise(sum=sum(CPUE)) %>%   # sum CPUE within Taxon
    spread(Taxon,sum,fill=0) ->    # "spread" taxon names across table header
    Taxon.CPUE
  summary(Taxon.CPUE)
  nrow(Taxon.CPUE)                 # 11,753 records

# latitude, longitude, etc 
  GOA_data %>%
    select(c(HAULJOIN,START_TIME,START_LONGITUDE,START_LATITUDE,BOTTOM_DEPTH,GEAR_TEMPERATURE,year)) %>%
    distinct() ->                  # choose distinct cases of the selected variables
    Location
  nrow(Location)                   # 11,753 sample locations

# match location and Taxon.CPUE data  
  Location.Taxon <- left_join(Location,Taxon.CPUE,by="HAULJOIN") 

# Add Taxon.presence (presence/absence = 1/0)
  # first create columns to store values
  num.records <- nrow(Location.Taxon)
  Taxon.presence <- as.data.frame(cbind(rep(0,num.records),rep(0,num.records),rep(0,num.records)))
  colnames(Taxon.presence) <- c("Coral.presence","SeaPenWhip.presence","Sponge.presence")
  Location.Taxon <- cbind(Location.Taxon[,-11],Taxon.presence)   # bind presence to location and CPUE
  for (i in 1:num.records) {                                     # determine presence/absence
    if(Location.Taxon$Coral[i]>0) Location.Taxon$Coral.presence[i] <- 1  
    if(Location.Taxon$SeaPenWhip[i]>0) Location.Taxon$SeaPenWhip.presence[i] <- 1 
    if(Location.Taxon$Sponge[i]>0) Location.Taxon$Sponge.presence[i] <- 1 
  }

# check the data some more
  summary(Location.Taxon)
  # 1,918 records with no temperature data
  # all other fields have data
# check years without temperature data
  Location.Taxon %>%
     filter(is.na(GEAR_TEMPERATURE)) %>%
    count(year) ->
    count.temp
  Location.Taxon %>%
    count(year) ->
    count.rec
  left_join(count.rec,count.temp,by="year")
# all years have some temperature data but less common for three earliest years (1984, 1987, 1990)
```

### read the (correctly projected) rasters and then stack them
```{r, cache=TRUE, echo=FALSE}

# set projection
newproj<-"+proj=aea +lat_1=50 +lat_2=58.5 +lat_0=45 +lon_0=-126 +x_0=1000000 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

# read GOA rasters from files
  variable.names <-  c("lon","lat","bathy","speed","btemp","color","slope","tmax")
  nvars <- length(variable.names)
  for (j in 1:nvars) {
   temp1<-raster(paste(projectionrasters.GOA.path,variable.names[j],sep=""))  # load the raster
   assign(paste(variable.names[j]),temp1)                                     # assign raster name
   }
  
raster.stack.GOA<-stack(lon,lat,bathy,speed,btemp,color,slope,tmax) # stack the rasters
names(raster.stack.GOA) <- variable.names                           # name the rasters 
plot(raster.stack.GOA)
# plot individually
plot(bathy)
plot(speed)
plot(btemp)
plot(color)
plot(slope)
plot(tmax)  # tmax is high outside Dixon Entrance and may account for the high inferred coral probability in the absence of any trawl survey samples
```

### match raster layers to the DSCS trawl data
```{r, cache=TRUE, echo=FALSE}
# all trawl survey locations, transform to a shape object, check overlay, extract raster values and bind to original data
Location <- data.frame(lon=Location.Taxon$START_LONGITUDE,lat=Location.Taxon$START_LATITUDE) # select locations
Location.GOA <- SpatialPoints(Location,proj4string=CRS("+proj=longlat +datum=WGS84"))     # create a list of the points
Location.GOA.projected <- spTransform(Location.GOA,crs(newproj))                    # add projection
plot(raster.stack.GOA[[3]])                                                         # plot projection
plot(Location.GOA.projected,add=TRUE)                                               # check overlay
GOA.extract<-raster::extract(raster.stack.GOA,Location.GOA.projected)               # extract raster values
mydata <- cbind(Location.Taxon,GOA.extract)
nrow(mydata)   # 11,753 records

# check for NA values (values that, for example, lie outside the GOA raster layers)
mydata %>%
  filter(is.na(speed)) ->
  mydata.na
nrow(mydata.na)   # 853 records
summary(mydata.na$START_LONGITUDE)
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# -170.0  -167.9  -166.4  -166.5  -165.5  -132.7
# most are in western Gulf of Alaska ==> outside the GOA raster

# remove these samples
mydata %>%
  filter(!is.na(speed)) ->
  mydata
nrow(mydata)   # 10,900 records
summary(mydata)   # only GEAR_TEMPERATURE WITH NA'S (AS EXPECTED)

# and last but not least
# drop records for 1984 and 1987 surveys (Chris questioned the accuracy of the location data)
mydata %>%
  filter(year >= 1990) ->
  mydata
nrow(mydata)
# 9,318 records
# which is the final version of mydata ...

# write final version
write.csv(mydata,paste(results.path,"Location.DSCS.trawl.csv",sep=""),row.names = FALSE)
```

### conform the raster and data variable names before running the models
```{r, cache=TRUE, echo=FALSE}
# trawl DSCS data
  mydata %>%
    select(-bathy) %>%              # drop bathy from raster layer 
    rename(depth = BOTTOM_DEPTH) -> # and instead use measured depth
    mydata

# and change the raster layers names to match
   names(raster.stack.GOA) <- c("lon","lat","depth","speed","btemp","color","slope","tmax") 
```

### separate train and test data sets
```{r, cache=TRUE, echo=FALSE}

# generate random uniform variables [0,1]
num.records <- nrow(mydata)
random.var <- runif(num.records)
hist(random.var)   # check random variables, looks good
length(random.var) # 9,318 records

# append to mydata
mydata <- cbind(mydata,random.var)

# fraction of samples allocated to training set
frac.test <- 0.8
mydata %>%
  filter(random.var <= frac.test) ->
  train
mydata %>%
  filter(random.var > frac.test) ->
  test
nrow(train)+nrow(test)   # check split, adds up to 9,318 ...

# choose non-zero values
train %>% 
  filter(Coral.presence==1) ->
  trainc
test %>% 
  filter(Coral.presence==1) ->
  testc 

# change variable names to match code
 train$corals_present <- train$Coral.presence
 test$corals_present <- test$Coral.presence
 
 # change variable names to match code
 trainc$Coral_generic <- trainc$Coral
 testc$Coral_generic <- testc$Coral


```

### analysis code (GAM, etc) from Chris Rooper
#### the next two chunks, if run under 'Knit', fail to plot the prediction rasters correctly
### METHOD 1 - DELTA GAM (GAM FOR PRESENCE-ABSENCE, SET THRESHOLD, GAM FOR POSITIVE OCCURRENCES
#### coralsGAMpa
```{r, cache=TRUE, echo=FALSE, eval=TRUE}
#fullmodel
 
# notes on variables:
#   use trawl survey depth (depth)
#   use temperature from raster (long-term average) (btemp)
# here is the presence-absence GAM
 
corals.DG.pa<-gam(corals_present~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4)+s(tmax,k=4)+s(speed,k=4)+s(slope,k=4)+s(color,k=4),data=train,family=binomial)
summary(corals.DG.pa)
# slope and color are not statistically significant

#reduced model (drop non-significant variables)
corals.DG.pa<-gam(corals_present~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4)+s(speed,k=4),data=train,family=binomial)
summary(corals.DG.pa)
gam.check(corals.DG.pa)
#plot(corals.DG.pa)

#test the predictions on the training data
train_predict<-predict.gam(corals.DG.pa,train,type="response")
train.auc_data<-data.frame(cbind(seq(1,length(train_predict),1),train[,"corals_present"],train_predict))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
# subset for values (i.e., remove NA)
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=6)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)   # observed 1 often predicted as 0
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3],method="spearman")

#test the predictions on the test data
test_predict<-predict.gam(corals.DG.pa,test,type="response")
test.auc_data<-data.frame(cbind(seq(1,length(test_predict),1),test[,"corals_present"],test_predict))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3],method="spearman")

corals.DG.pa.raster<-predict(raster.stack.GOA, corals.DG.pa,filename=paste(predictionrasters.path,"coralsGAMpa",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)
plot(corals.DG.pa.raster,main="Coral Generalized Additive Model - PA")
#plot(akland,col=1,add=TRUE)
#plot(canada_land,col=1,add=TRUE)
```

### GAM model
#### coralsGAMcpue
```{r, cache=TRUE, echo=FALSE, eval=TRUE}
############GAM Models for Coral abundance

#fullmodel
corals.DG.cpue<-gam((Coral_generic)^.25~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4)+s(tmax,k=4)+s(speed,k=4)+s(slope,k=4)+s(color,k=4),data=trainc,family=gaussian)
summary(corals.DG.cpue)

#reduced model
corals.DG.cpue<-gam((Coral_generic)^.25~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4)+s(slope,k=4),data=trainc,family=gaussian)
colnames(train)
summary(corals.DG.cpue)
gam.check(corals.DG.cpue)
#plot(corals.DG.cpue,scale=0,residuals=TRUE)

#test the predictions on the training data
observed.train<-(trainc$Coral_generic)^.25
predicted.train<-predict(corals.DG.cpue,trainc,type="response")
pred.train<-lm(observed.train~predicted.train)
summary(pred.train)
plot(observed.train,predicted.train)
abline(pred.train)

#test the predictions on the test data
observed.test<-(testc$Coral_generic)^.25
predicted.test<-predict(corals.DG.cpue,testc,type="response")
pred.test<-lm(observed.test~predicted.test)
summary(pred.test)
plot(observed.test,predicted.test)
abline(pred.test)

# make a prediction raster
# probability of presence (GLM 0/1) and abundance raster
corals.DG.cpue.raster<-predict(raster.stack.GOA, corals.DG.cpue,filename=paste(predictionrasters.path,"coralsGAMcpue",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)

plot(corals.DG.cpue.raster, main="Coral GAM - PA, GAM for CPUE, Threshold = 0.05")
#plot(akland,col=1,add=TRUE)
#plot(canada_land,col=1,add=TRUE)

#####OVERALL R2 GAM######
train.observed<-(train$Coral)^0.25
length(train.observed)
train.predicted<-predict(corals.DG.pa,train,type="response")
train.predicted[train.predicted>=train.threshold]<-1
train.predicted[train.predicted<train.threshold]<-0
train.predicted.cpue<-predict(corals.DG.cpue,train,type="response")
train.predicted.cpue<-train.predicted.cpue*train.predicted
pred.train<-lm(train.predicted.cpue~train.observed)
summary(pred.train)
plot(train.observed,train.predicted.cpue)
abline(pred.train)
mse_coral<-sum(pred.train$residuals^2)/pred.train$df.residual
mse_coral

test.observed<-(test$Coral)^0.25
test.predicted<-predict(corals.DG.pa,test,type="response")
test.predicted[test.predicted>=train.threshold]<-1
test.predicted[test.predicted<train.threshold]<-0
test.predicted.cpue<-predict(corals.DG.cpue,test,type="response")
test.predicted.cpue<-test.predicted.cpue*test.predicted
pred.test<-lm(test.predicted.cpue~test.observed)
summary(pred.test)
plot(test.observed,test.predicted.cpue)
abline(pred.test)
mse_coral<-sum(pred.test$residuals^2)/pred.test$df.residual
mse_coral
```

### METHOD 2 - Boosted Regression Tree
#### coralsBRTpa
#### coralsBRTcpue
```{r, cache=TRUE, echo=FALSE, eval=TRUE}
corals.BRT.pa<-gbm.step(data=train, gbm.x = c(5,14:20), gbm.y = 11, family = "bernoulli", tree.complexity = 5,learning.rate = 0.005, bag.fraction = 0.5)
gbm.plot(corals.BRT.pa)
gbm.plot.fits(corals.BRT.pa)
summary(corals.BRT.pa)

#test the predictions on the training data
train_predict<-predict.gbm(corals.BRT.pa,train, n.trees=corals.BRT.pa$gbm.call$best.trees, type="response")
train.auc_data<-data.frame(cbind(seq(1,length(train_predict),1),train[,"corals_present"],train_predict))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=2)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3])

#test the predictions on the test data
test_predict<-predict.gbm(corals.BRT.pa,test, n.trees=corals.BRT.pa$gbm.call$best.trees, type="response")
test.auc_data<-data.frame(cbind(seq(1,length(test_predict),1),test[,"corals_present"],test_predict))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3])

#Predict to Raster
corals.BRT.pa.raster<-predict(raster.stack.GOA, corals.BRT.pa, n.trees=corals.BRT.pa$gbm.call$best.trees, type="response",filename=paste(predictionrasters.path,"coralsBRTpa",sep=""), na.rm=TRUE,overwrite=TRUE,progress="text",newdata.guaranteed=TRUE)
plot(corals.BRT.pa.raster,main="Coral Boosted Regression Tree - PA")
#plot(akland,col=1,add=TRUE)
#plot(canada_land,col=1,add=TRUE)

#############BRT Models for Coral abundance
# change variable names to match code
train$Coral_generic <- train$Coral
test$Coral_generic <- test$Coral

train$corals.fourth<-(train$Coral_generic)^0.25
test$corals.fourth<-(test$Coral_generic)^0.25
corals.BRT<-gbm.step(data=train, gbm.x = c(5,14:20), gbm.y = 24, family = "gaussian", tree.complexity = 5,learning.rate = 0.01, bag.fraction = 0.5)
gbm.plot(corals.BRT)
gbm.plot.fits(corals.BRT)
summary(corals.BRT)

#Test on training data
train_predict<-predict.gbm(corals.BRT,train, n.trees=corals.BRT$gbm.call$best.trees, type="response")
plot(train_predict~train$corals.fourth)
brt.pred.train<-lm(train_predict~train$corals.fourth)
summary(brt.pred.train)
abline(brt.pred.train)
mse_coral<-sum(brt.pred.train$residuals^2)/brt.pred.train$df.residual
mse_coral

#Test on test data
test_predict<-predict.gbm(corals.BRT,test, n.trees=corals.BRT$gbm.call$best.trees, type="response")
plot(test_predict~test$corals.fourth)
brt.pred.test<-lm(test_predict~test$corals.fourth)
summary(brt.pred.test)
abline(brt.pred.test)
mse_coral<-sum(brt.pred.test$residuals^2)/brt.pred.test$df.residual
mse_coral

#Predict to Raster
corals.BRT.cpue.raster<-predict(raster.stack.GOA, corals.BRT, n.trees=corals.BRT$gbm.call$best.trees, type="response",filename=paste(predictionrasters.path,"coralsBRTcpue",sep=""), na.rm=TRUE,overwrite=TRUE,progress="text",newdata.guaranteed=TRUE)
plot(corals.BRT.cpue.raster,main="Coral Boosted Regression Tree - PA")
#plot(akland,col=1,add=TRUE)
#plot(canada_land,col=1,add=TRUE)
```

### METHOD 3 - DELTA GLM METHOD
#### coralsGLMpa
#### coralsGLMcpue
```{r, cache=TRUE, echo=FALSE, eval=TRUE}
#fullmodel
corals.GLM.pa<-glm(corals_present~lon+lat+depth+btemp+slope+tmax+speed+color+I(lon^2)++I(lat^2)+I(depth^2)+I(slope^2)+I(btemp^2)+I(tmax^2)+I(speed^2)+I(color^2),data=train,family=binomial)
summary(corals.GLM.pa)

#reduced model
corals.GLM.pa<-glm(corals_present~lon+lat+tmax+speed+color+I(tmax^2)+I(speed^2),data=train,family=binomial)
summary(corals.GLM.pa)
#plot(corals.GLM.pa)

#test the predictions on the training data
train_predict<-predict.glm(corals.GLM.pa,train,type="response")
train.auc_data<-data.frame(cbind(seq(1,length(train_predict),1),train[,"corals_present"],train_predict))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=6)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3],method="spearman")

#test the predictions on the test data
test_predict<-predict.glm(corals.GLM.pa,test,type="response")
test.auc_data<-data.frame(cbind(seq(1,length(test_predict),1),test[,"corals_present"],test_predict))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3])

corals.GLM.pa.raster<-predict(raster.stack.GOA, corals.GLM.pa,filename=paste(predictionrasters.path,"coralsGLMpa",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)
plot(corals.GLM.pa.raster,main="Coral Generalized Linear Model - PA")
#plot(akland,col=1,add=TRUE)
#plot(canada_land,col=1,add=TRUE)

#############GLM Models for Coral abundance
#fullmodel
corals.GLM.cpue<-glm((Coral_generic)^.25~lon+lat+depth+btemp+slope+tmax+speed+color+I(lon^2)+I(lat^2)+I(depth^2)+I(slope^2)+I(btemp^2)+I(tmax^2)+I(speed^2)+I(color^2),data=trainc,family=gaussian)
summary(corals.GLM.cpue)

#reduced model
corals.GLM.cpue<-glm((Coral_generic)^.25~lon+lat+depth+btemp+slope+color+I(lat^2)+I(btemp^2)+I(color^2),data=trainc,family=gaussian)
summary(corals.GLM.cpue)
#plot(corals.GLM.cpue,scale=0,residuals=TRUE)

#test the predictions on the training data
observed.train.GLM<-(trainc$Coral_generic)^.25
predicted.train.GLM<-predict(corals.GLM.cpue,trainc,type="response")
pred.train.GLM<-lm(observed.train.GLM~predicted.train.GLM)
summary(pred.train.GLM)
plot(observed.train.GLM,predicted.train.GLM)
abline(pred.train.GLM)

#test the predictions on the test data
observed.test.GLM<-(testc$Coral_generic)^.25
predicted.test.GLM<-predict(corals.GLM.cpue,testc,type="response")
pred.test.GLM<-lm(observed.test.GLM~predicted.test.GLM)
summary(pred.test.GLM)
plot(observed.test.GLM,predicted.test.GLM)
abline(pred.test.GLM)

temp_raster<-cut(corals.GLM.pa.raster,breaks=c(train.threshold,1))
corals.GLM.cpue.raster<-predict(raster.stack.GOA, corals.GLM.cpue,filename=paste(predictionrasters.path,"coralsGLMcpue",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)

plot(corals.GLM.cpue.raster, main="Coral GAM - PA, GAM for CPUE, Threshold = 0.05")
#plot(akland,col=1,add=TRUE)
#plot(canada_land,col=1,add=TRUE)

#####OVERALL R2 GLM######
train.observed<-(train$Coral)^0.25
train.predicted<-predict(corals.GLM.pa,train,type="response")
train.predicted[train.predicted>=train.threshold]<-1
train.predicted[train.predicted<train.threshold]<-0
train.predicted.cpue<-predict(corals.GLM.cpue,train,type="response")
train.predicted.cpue<-train.predicted.cpue*train.predicted
pred.train<-lm(train.predicted.cpue~train.observed)
summary(pred.train)
plot(train.observed,train.predicted.cpue)
abline(pred.train)
mse_coral<-sum(pred.train$residuals^2)/pred.train$df.residual
mse_coral

test.observed<-(test$Coral)^0.25
test.predicted<-predict(corals.GLM.pa,test,type="response")
test.predicted[test.predicted>=train.threshold]<-1
test.predicted[test.predicted<train.threshold]<-0
test.predicted.cpue<-predict(corals.GLM.cpue,test,type="response")
test.predicted.cpue<-test.predicted.cpue*test.predicted
pred.test<-lm(test.predicted.cpue~test.observed)
summary(pred.test)
plot(test.observed,test.predicted.cpue)
abline(pred.test)
mse_coral<-sum(pred.test$residuals^2)/pred.test$df.residual
mse_coral
```

### METHOD 4 - RANDOM FOREST METHOD
#### coralsRFpa
```{r, cache=TRUE, echo=FALSE, eval=TRUE}
corals.RF.pa<-randomForest(as.factor(corals_present)~lat+lon+depth+slope+btemp+speed+tmax+color,data=train, ntree=1000,importance=TRUE, na.action=na.omit)
print(corals.RF.pa)
round(importance(corals.RF.pa), 2)

train_predict_RF<-predict(corals.RF.pa,train,type="prob")[,2]
train.auc_data<-data.frame(cbind(seq(1,length(train_predict_RF),1),train[,"corals_present"],train_predict_RF))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=2)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3])


#Test on test data
test_predict_RF<-predict(corals.RF.pa,test,type="prob")[,2]
test.auc_data<-data.frame(cbind(seq(1,length(test_predict_RF),1),test[,"corals_present"],test_predict_RF))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)  # random forest overfits the data
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3])


#Predict to Raster
corals.RF.pa.raster<-predict(raster.stack.GOA, corals.RF.pa,type="prob",filename=paste(predictionrasters.path,"coralsRFpa",sep=""), na.rm=TRUE,overwrite=TRUE,progress="text",newdata.guaranteed=TRUE)
plot(1-corals.RF.pa.raster,main="Coral Boosted Regression Tree - PA")   # was the probability of the zero category
#plot(akland,col=1,add=TRUE)
#plot(canada_land,col=1,add=TRUE)

removeTmpFiles(h=.25)

```

### read the presence-absence rasters, stack them, and then compute their mean
```{r, cache=TRUE, echo=FALSE}

# read prediction rasters from files
  variable.names <-  c("coralsBRTpa","coralsGAMpa","coralsGLMpa","coralsRFpa")
  nvars <- length(variable.names)
  for (j in 1:nvars) {
   temp1<-raster(paste(predictionrasters.path,variable.names[j],sep=""))  # load the raster
   assign(paste(variable.names[j]),temp1)                                 # assign raster name
   }

# convert coralsRFpa from probability of absence to probability of presence  
coralsRFpa <- 1 - coralsRFpa    

compareRaster(coralsBRTpa,coralsGAMpa,coralsGLMpa,coralsRFpa)

raster.stack.pa<-stack(coralsGAMpa,coralsGLMpa,coralsBRTpa,coralsRFpa) # stack the rasters
names(raster.stack.pa) <- c("GAM","GLM","BRT","RF")                           # name the rasters 
plot(raster.stack.pa)

# compute, plot and write to a file the mean probability of presence
pa.mean <- mean(raster.stack.pa)
plot(pa.mean)
plot(Location.GOA.projected,add=TRUE, cex=0.0001, pch=1)   # overlay sample locations  
writeRaster(pa.mean,filename=paste(predictionrasters.path,"pamean",sep=""),overwrite=T)
```
