---
title: "read trawl data"
author: "Mike Sigler"
date: "4/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### read trawl data.Rmd
#### analyze Gulf of Alaska trawl survey data for deep sea coral and sponge

### read libraries
```{r,cache=TRUE,echo=FALSE}

## load libraries
library(mgcv)
library(PresenceAbsence)
library(gbm)
library(dismo)
library(raster)
library(rgdal)
library(maptools)
library(gstat)
library(rgeos)
library(proj4)
library(sp)
library(maptools)
library(maps)
library(xtable)
library(randomForest)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(PerformanceAnalytics)
library(ggfortify)
library(lubridate)
```

### set up directories
```{r, cache=TRUE, echo=FALSE}
#rm(list = ls())
# set project directories
data.path     <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Trawl data\\"
GOArasters.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Variables_GOA_1km\\"
AIrasters.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Variables_AI_1km_1000m\\"
results.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Results\\"
predictionrasters.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\Results\\PredictionRasters\\"
projectionrasters.GOA.path <- "D:\\Documents\\Alaska-Coral-and-Sponge-Model\\ProjectionRastersGOA\\"
```

### read trawl data
```{r, cache=TRUE, echo=FALSE}
# read trawl data
GOA_data <-as_tibble(read.csv(paste(data.path,"GOA_Trawl_data.csv",sep="")))
GOA_data$year <- year(mdy_hms(GOA_data$START_TIME))

# some data checks
# cross tabulate
GOA_data %>%
  count(Taxon)   # looks good
# locations
GOA_data %>%
  ggplot(aes(x=START_LONGITUDE)) + geom_histogram()  # looks good
GOA_data %>%
  ggplot(aes(x=START_LATITUDE)) + geom_histogram()   # looks good
GOA_data %>%
  count(year)   # looks good

# Group data by haul
# Taxon CPUE
  GOA_data %>% 
    group_by(HAULJOIN,Taxon) %>%
    summarise(sum=sum(CPUE)) %>%   # sum CPUE within Taxon
    spread(Taxon,sum,fill=0) ->    # "spread" taxon names across table header
    Taxon.CPUE
  summary(Taxon.CPUE)
  nrow(Taxon.CPUE)                 # 11,753 records

# latitude, longitude, etc 
  GOA_data %>%
    select(c(HAULJOIN,START_TIME,START_LONGITUDE,START_LATITUDE,BOTTOM_DEPTH,GEAR_TEMPERATURE,year)) %>%
    distinct() ->                  # choose distinct cases of the selected variables
    Location
  nrow(Location)                   # 11,753 camera drops

# match location and Taxon.CPUE data  
  Location.Taxon <- left_join(Location,Taxon.CPUE,by="HAULJOIN") 

# Add Taxon.presence (presence/absence = 1/0)
  # first create columns to store values
  num.records <- nrow(Location.Taxon)
  Taxon.presence <- as.data.frame(cbind(rep(0,num.records),rep(0,num.records),rep(0,num.records)))
  colnames(Taxon.presence) <- c("Coral.presence","SeaPenWhip.presence","Sponge.presence")
  Location.Taxon <- cbind(Location.Taxon[,-11],Taxon.presence)   # bind presence to location and CPUE
  for (i in 1:num.records) {                                     # determine presence/absence
    if(Location.Taxon$Coral[i]>0) Location.Taxon$Coral.presence[i] <- 1  
    if(Location.Taxon$SeaPenWhip[i]>0) Location.Taxon$SeaPenWhip.presence[i] <- 1 
    if(Location.Taxon$Sponge[i]>0) Location.Taxon$Sponge.presence[i] <- 1 
  }

# check the data some more
  summary(Location.Taxon)
  # 1,918 records with no temperature data
  # all other fields have data
# check years without temperature data
  Location.Taxon %>%
     filter(is.na(GEAR_TEMPERATURE)) %>%
    count(year) ->
    count.temp
  Location.Taxon %>%
    count(year) ->
    count.rec
  left_join(count.rec,count.temp,by="year")
# all years have some temperature data but less common for three earliest years (1984, 1987, 1990)
```

### read the (correctly projected) rasters and then stack them
```{r, cache=TRUE, echo=FALSE,eval=FALSE}

# set projection
newproj<-"+proj=aea +lat_1=50 +lat_2=58.5 +lat_0=45 +lon_0=-126 +x_0=1000000 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

# read GOA rasters from files
  variable.names <-  c("lon","lat","bathy","speed","btemp","color","slope","tmax")
  nvars <- length(variable.names)
  for (j in 1:nvars) {
   temp1<-raster(paste(projectionrasters.GOA.path,variable.names[j],sep=""))  # load the raster
   assign(paste(variable.names[j]),temp1)                                     # assign raster name
   }
  
raster.stack.GOA<-stack(lon,lat,bathy,speed,btemp,color,slope,tmax) # stack the rasters
names(raster.stack.GOA) <- variable.names                           # name the rasters 
plot(raster.stack.GOA)                                              # plot each raster
```

### match raster layers to the DSCS trawl data
```{r, cache=TRUE, echo=FALSE}
# all trawl survey locations, transform to a shape object, check overlay, extract raster values and bind to original data
Location <- data.frame(lon=Location.Taxon$START_LONGITUDE,lat=Location.Taxon$START_LATITUDE) # select locations
Location.GOA <- SpatialPoints(Location,proj4string=CRS("+proj=longlat +datum=WGS84"))     # create a list of the points
Location.GOA.projected <- spTransform(Location.GOA,crs(newproj))                    # add projection
plot(raster.stack.GOA[[3]])                                                         # plot projection
plot(Location.GOA.projected,add=TRUE)                                               # check overlay
GOA.extract<-raster::extract(raster.stack.GOA,Location.GOA.projected)               # extract raster values
mydata <- cbind(Location.Taxon,GOA.extract)
nrow(mydata)   # 11,753 records

# check for NA values (values that, for example, lie outside the GOA raster layers)
mydata %>%
  filter(is.na(speed)) ->
  mydata.na
nrow(mydata.na)   # 853 records
summary(mydata.na$START_LONGITUDE)
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# -170.0  -167.9  -166.4  -166.5  -165.5  -132.7
# most are in western Gulf of Alaska

# remove columns with NA values (for bathy, etc)
#  the resulting table will be tested against the Aleutians raster layers
mydata.na %>%
  select(-c(lon:tmax)) ->
  test.GOA

# match the Gulf of Alaska NA values to the Aleutians rasters, which extends east of 170 deg W
Location.testGOA <- data.frame(lon=test.GOA$START_LONGITUDE,lat=test.GOA$START_LATITUDE)  # identify data
Location.testGOA <- SpatialPoints(Location.testGOA,proj4string=CRS("+proj=longlat +datum=WGS84")) # create a list of the points
Location.testGOA.projected <- spTransform(Location.testGOA,crs(newproj))    # add projection
plot(raster.stack.AI[[3]])                                                  # plot projection
plot(Location.testGOA.projected,add=TRUE)                                   # check overlay
testGOA.extract<-raster::extract(raster.stack.AI,Location.testGOA.projected)# extract raster values
Location.DSCS.testGOA<-cbind(test.GOA,testGOA.extract)

Location.DSCS.testGOA %>%
  filter(is.na(speed)) ->       # check for NA values 
  test2.GOA
nrow(Location.DSCS.testGOA)        # 853 records (outside GOA raster)
nrow(test2.GOA)                    #  59 records ==> some records remain outside AI raster 
max(test2.GOA$START_LONGITUDE); min(test2.GOA$START_LONGITUDE)    # no obvious rhyme or reason to the excluded points  
# [1] -132.6795
# [1] -163.2772
max(test2.GOA$BOTTOM_DEPTH); min(test2.GOA$BOTTOM_DEPTH)             
# 787
# [1] 33
max(test2.GOA$START_LATITUDE); min(test2.GOA$START_LATITUDE)  
# [1] 58.116
# [1] 54.48439

# add indicator for sample locations outside the GOA rasters
Location.DSCS.testGOA$inside.GOA.raster <- 0

# replace records outside the GOA raster, now with AI information (in most cases), in mydata
mydata %>%
  filter(!is.na(speed)) ->       # remove samples outside the GOA raster
  mydata.revised
mydata.revised$inside.GOA.raster <- 1  # add indicator for records inside the GOA raster
nrow(mydata.revised)                # 10,900 records
nrow(Location.DSCS.testGOA)         # 853 records
names(mydata.revised)
names(Location.DSCS.testGOA)
mydata.revised <- rbind(mydata.revised,Location.DSCS.testGOA) # revised GOA table
nrow(mydata.revised)                # 11,753 records

mydata.revised %>%
  filter(is.na(bathy)) ->       # check for remaining NA values, should be 59 records 
  test2.GOA
nrow(test2.GOA)                     # 59 records, which matches the number of locations outside both
                                    #   the AI and GOA rasters

mydata.revised %>%              # and remove those records
  filter(!is.na(bathy)) ->
  mydata
nrow(mydata)                    # 11,694 records

# and last but not least
# drop records for 1984 and 1987 surveys (Chris questioned the accuracy of the location data)
mydata %>%
  filter(year >= 1990) ->
  mydata
nrow(mydata)
# 9,992 records
# which is the final version of mydata ...

# write final version
write.csv(mydata,paste(results.path,"Location.DSCS.trawl.csv",sep=""),row.names = FALSE)
```



### read the matched data (if you do not make the raster stacks)
```{r, cache=TRUE, echo=FALSE,eval=FALSE}
#  if you do not want to rerun all of the previous code, which takes ~2 hours
#    run the next (commented) step
  mydata <- read.csv(paste(results.path,"Location.DSCS.trawl.csv",sep=""))

# alternately, restack GOA rasters, which are used for predictions (takes ~3 minutes)
# set projection
newproj<-"+proj=aea +lat_1=50 +lat_2=58.5 +lat_0=45 +lon_0=-126 +x_0=1000000 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"

# import rasters
variables <- c("Bathy","Bcurrent","Btemp","Color","Slope","Tmax")   # raster (variable) names
nvar <- length(variables)                                           # number of variables
areas <- c("AI","EBS","GOA")                                        # number of areas

# Gulf of Alaska
for (j in 1:nvar) {
  temp1<-raster(paste(GOArasters.path,variables[j],sep=""))                      # load the raster
  temp2<-projectRaster(temp1,crs=newproj,progress="text")                        # convert the projection
  assign(paste(variables[j],".",areas[3],".","projected",sep=""),temp2)          # assign raster name
}
raster.stack.GOA<-stack(Bathy.GOA.projected,Bcurrent.GOA.projected,Btemp.GOA.projected, # stack the rasters
                       Color.GOA.projected,Slope.GOA.projected,Tmax.GOA.projected)
names(raster.stack.GOA) <- variables                                             # name the rasters 
plot(raster.stack.GOA)                                                           # plot each raster

# Create lat and lon rasters (needed for predictions) and stack with other rasters
 lon.GOA.projected <- init(Bathy.GOA.projected,v='x')
 lat.GOA.projected<-init(Bathy.GOA.projected,v='y')
 raster.stack.GOA <- stack(lon.GOA.projected,lat.GOA.projected,raster.stack.GOA)
 variable.names <-  c("lon","lat","bathy","speed","btemp","color","slope","tmax")
 names(raster.stack.GOA) <-  variable.names 
 plot(raster.stack.GOA)
# should be good to go to complete analyses without having to rerun AI rasters
```



### conform the raster and data variable names before running the models
```{r, cache=TRUE, echo=FALSE}
# trawl DSCS data
  mydata %>%
    select(-bathy) %>%              # drop bathy from raster layer 
    rename(depth = BOTTOM_DEPTH) -> # and instead use measured depth
    mydata
  
# and change the raster layers names to match
   names(raster.stack.GOA) <- c("lon","lat","depth","speed","btemp","color","slope","tmax") 
   plot(raster.stack.GOA)
```

#### analysis code (GAM, etc) from Chris Rooper
### separate train and test data sets
```{r, cache=TRUE, echo=FALSE}

# exclude sample locations outside GOA rasters (necessary for prediction)
mydata %>%
  filter(inside.GOA.raster==1) ->
  mydata
nrow(mydata)   # 9,318 records 
summary(mydata)  # 652 NA for gear temperature, which is not used in the calculations
                 # instead average temperature (btemp) is used in calculations

# generate random uniform variables [0,1]
num.records <- nrow(mydata)
random.var <- runif(num.records)
hist(random.var)   # check random variables, looks good
length(random.var) # 9,318 records

# append to mydata
mydata <- cbind(mydata,random.var)

# fraction of samples allocated to training set
frac.test <- 0.8
mydata %>%
  filter(random.var <= frac.test) ->
  train
mydata %>%
  filter(random.var > frac.test) ->
  test
nrow(train)+nrow(test)   # check split, adds up to 9,318 ...

# choose non-zero values
train %>% 
  filter(Coral.presence==1) ->
  trainc
test %>% 
  filter(Coral.presence==1) ->
  testc 
```


### METHOD 1 - DELTA GAM (GAM FOR PRESENCE-ABSENCE, SET THRESHOLD, GAM FOR POSITIVE OCCURRENCES
```{r, cache=TRUE, echo=FALSE}
#fullmodel
# use Chris's names
 train$corals_present <- train$Coral.presence
 test$corals_present <- test$Coral.presence
 
# notes on variables:
#   use trawl survey depth (depth)
#   use temperature from raster (long-term average) (btemp)
# here is the presence-absence GAM
 
corals.DG.pa<-gam(corals_present~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4)+s(tmax,k=4)+s(speed,k=4)+s(slope,k=4)+s(color,k=4),data=train,family=binomial)
summary(corals.DG.pa)
# slope and color are not statistically significant

#reduced model (drop non-significant variables)
corals.DG.pa<-gam(corals_present~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4)+s(tmax,k=4)+s(speed,k=4),data=train,family=binomial)
summary(corals.DG.pa)
gam.check(corals.DG.pa)
#plot(corals.DG.pa)

#test the predictions on the training data
train_predict<-predict.gam(corals.DG.pa,train,type="response")
train.auc_data<-data.frame(cbind(seq(1,length(train_predict),1),train[,"corals_present"],train_predict))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
# subset for values (i.e., remove NA)
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=6)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)   # observed 1 often predicted as 0
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3],method="spearman")

#test the predictions on the test data
test_predict<-predict.gam(corals.DG.pa,test,type="response")
test.auc_data<-data.frame(cbind(seq(1,length(test_predict),1),test[,"corals_present"],test_predict))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3],method="spearman")

corals.DG.pa.raster<-predict(raster.stack.GOA, corals.DG.pa,filename=paste(predictionrasters.path,"coralsGAMpa",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)
plot(corals.DG.pa.raster,main="Coral Generalized Additive Model - PA")
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)
# Chris, drop this next line, right? (related to dropped code in GAM section)
#corals.presence.mask<-mask(corals.DG.pa.raster,corals.habitat.mask)
#writeRaster(corals.presence.mask,filename=paste(predictionrasters.path,"coralsGAMpa",sep=""),overwrite=TRUE)
```


### GAM model
```{r, cache=TRUE, echo=FALSE}
############GAM Models for Coral abundance
# change variable names to match code

 trainc$Coral_generic <- trainc$Coral
 testc$Coral_generic <- testc$Coral

#fullmodel
 # there was an error here, which Chris diagnosed
corals.DG.cpue<-gam((Coral_generic)^.25~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4)+s(tmax,k=4)+s(speed,k=4)+s(slope,k=4)+s(color,k=4),data=trainc,family=gaussian)
summary(corals.DG.cpue)

#reduced model
corals.DG.cpue<-gam((Coral_generic)^.25~s(lon)+s(lat)+s(depth,k=4)+s(btemp,k=4),data=trainc,family=gaussian)
colnames(train)
summary(corals.DG.cpue)
gam.check(corals.DG.cpue)
#plot(corals.DG.cpue,scale=0,residuals=TRUE)

#test the predictions on the training data
observed.train<-(trainc$Coral_generic)^.25
predicted.train<-predict(corals.DG.cpue,trainc,type="response")
pred.train<-lm(observed.train~predicted.train)
summary(pred.train)
plot(observed.train,predicted.train)
abline(pred.train)

#test the predictions on the test data
observed.test<-(testc$Coral_generic)^.25
predicted.test<-predict(corals.DG.cpue,testc,type="response")
pred.test<-lm(observed.test~predicted.test)
summary(pred.test)
plot(observed.test,predicted.test)
abline(pred.test)

# make a prediction raster
# probability of presence (GLM 0/1) and abundance raster
corals.DG.cpue.raster<-predict(raster.stack.GOA, corals.DG.cpue,filename=paste(predictionrasters.path,"coralsGAMcpue",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)

plot(corals.DG.cpue.raster, main="Coral GAM - PA, GAM for CPUE, Threshold = 0.05")
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)

#####OVERALL R2 GAM######
### the correction is here
#train.observed<-(train$Coral_generic)^0.25
train.observed<-(train$Coral)^0.25
length(train.observed)
train.predicted<-predict(corals.DG.pa,train,type="response")
train.predicted[train.predicted>=train.threshold]<-1
train.predicted[train.predicted<train.threshold]<-0
train.predicted.cpue<-predict(corals.DG.cpue,train,type="response")
train.predicted.cpue<-train.predicted.cpue*train.predicted
# Chris: fails here (occurs later too ... ) (not any more)
pred.train<-lm(train.predicted.cpue~train.observed)
summary(pred.train)
plot(train.observed,train.predicted.cpue)
abline(pred.train)
mse_coral<-sum(pred.train$residuals^2)/pred.train$df.residual
mse_coral

test.observed<-(test$Coral)^0.25
test.predicted<-predict(corals.DG.pa,test,type="response")
test.predicted[test.predicted>=train.threshold]<-1
test.predicted[test.predicted<train.threshold]<-0
test.predicted.cpue<-predict(corals.DG.cpue,test,type="response")
test.predicted.cpue<-test.predicted.cpue*test.predicted
# Chris, fails here
pred.test<-lm(test.predicted.cpue~test.observed)
summary(pred.test)
plot(test.observed,test.predicted.cpue)
abline(pred.test)
mse_coral<-sum(pred.test$residuals^2)/pred.test$df.residual
mse_coral
```


### METHOD 2 - Boosted Regression Tree
```{r, cache=TRUE, echo=FALSE}
####BRT PRESENCE ABSENCE MODEL######################################
# Chris, the next step fails (I revised the selected columns)
corals.BRT.pa<-gbm.step(data=train, gbm.x = c(5,14:20), gbm.y = 11, family = "bernoulli", tree.complexity = 5,learning.rate = 0.005, bag.fraction = 0.5)
gbm.plot(corals.BRT.pa)
gbm.plot.fits(corals.BRT.pa)
summary(corals.BRT.pa)

#test the predictions on the training data
train_predict<-predict.gbm(corals.BRT.pa,train, n.trees=corals.BRT.pa$gbm.call$best.trees, type="response")
train.auc_data<-data.frame(cbind(seq(1,length(train_predict),1),train[,"corals_present"],train_predict))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=2)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3])

#test the predictions on the test data
test_predict<-predict.gbm(corals.BRT.pa,test, n.trees=corals.BRT.pa$gbm.call$best.trees, type="response")
test.auc_data<-data.frame(cbind(seq(1,length(test_predict),1),test[,"corals_present"],test_predict))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3])



#Predict to Raster
corals.BRT.pa.raster<-predict(raster.stack.GOA, corals.BRT.pa, n.trees=corals.BRT.pa$gbm.call$best.trees, type="response",filename=paste(predictionrasters.path,"coralsBRTpa",sep=""), na.rm=TRUE,overwrite=TRUE,progress="text",newdata.guaranteed=TRUE)
plot(corals.BRT.pa.raster,main="Coral Boosted Regression Tree - PA")
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)

#############BRT Models for Coral abundance
train$corals.fourth<-(train$Coral_generic)^0.25
test$corals.fourth<-(test$Coral_generic)^0.25
corals.BRT<-gbm.step(data=train, gbm.x = c(8,29:30,32,34:36,38), gbm.y = 40, family = "gaussian", tree.complexity = 5,learning.rate = 0.01, bag.fraction = 0.5)
gbm.plot(corals.BRT)
gbm.plot.fits(corals.BRT)
summary(corals.BRT)

#Test on training data
train_predict<-predict.gbm(corals.BRT,train, n.trees=corals.BRT$gbm.call$best.trees, type="response")
plot(train_predict~train$corals.fourth)
brt.pred.train<-lm(train_predict~train$corals.fourth)
summary(brt.pred.train)
abline(brt.pred.train)
mse_coral<-sum(brt.pred.train$residuals^2)/brt.pred.train$df.residual
mse_coral

#Test on test data
test_predict<-predict.gbm(corals.BRT,test, n.trees=corals.BRT$gbm.call$best.trees, type="response")
plot(test_predict~test$corals.fourth)
brt.pred.test<-lm(test_predict~test$corals.fourth)
summary(brt.pred.test)
abline(brt.pred.test)
mse_coral<-sum(brt.pred.test$residuals^2)/brt.pred.test$df.residual
mse_coral

#Predict to Raster
corals.BRT.cpue.raster<-predict(GOA.stack, corals.BRT, n.trees=corals.BRT$gbm.call$best.trees, type="response",filename="U:/GOA Coral and Sponge Model/Results/PredictionRasters/coralsBRTcpue", na.rm=TRUE,overwrite=TRUE,progress="text",newdata.guaranteed=TRUE)
corals.BRT.cpue.raster<-mask(corals.BRT.cpue.raster,corals.habitat.mask,overwrite=TRUE)
writeRaster(corals.BRT.cpue.raster,filename="U:/GOA Coral and Sponge Model/Results/PredictionRasters/coralsBRTcpue",overwrite=TRUE)
plot(corals.BRT.cpue.raster,main="Coral Boosted Regression Tree - PA")
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)
```


### METHOD 3 - DELTA GLM METHOD
```{r, cache=TRUE, echo=FALSE}
#fullmodel
corals.GLM.pa<-glm(corals_present~lon+lat+depth+btemp+slope+tmax+speed+color+I(lon^2)++I(lat^2)+I(depth^2)+I(slope^2)+I(btemp^2)+I(tmax^2)+I(speed^2)+I(color^2),data=train,family=binomial)
summary(corals.GLM.pa)

#reduced model
corals.GLM.pa<-glm(corals_present~lon+lat+btemp+tmax+speed+color+I(lon^2)++I(lat^2)+I(speed^2)+I(color^2),data=train,family=binomial)
summary(corals.GLM.pa)
#plot(corals.GLM.pa)

#test the predictions on the training data
train_predict<-predict.glm(corals.GLM.pa,train,type="response")
train.auc_data<-data.frame(cbind(seq(1,length(train_predict),1),train[,"corals_present"],train_predict))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=6)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3],method="spearman")

#test the predictions on the test data
test_predict<-predict.glm(corals.GLM.pa,test,type="response")
test.auc_data<-data.frame(cbind(seq(1,length(test_predict),1),test[,"corals_present"],test_predict))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3])

corals.GLM.pa.raster<-predict(raster.stack.GOA, corals.GLM.pa,filename=paste(predictionrasters.path,"coralsGLMpa",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)
plot(corals.GLM.pa.raster,main="Coral Generalized Linear Model - PA")
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)

#############GLM Models for Coral abundance
#fullmodel
corals.GLM.cpue<-glm((Coral_generic)^.25~lon+lat+depth+btemp+slope+tmax+speed+color+I(lon^2)+I(lat^2)+I(depth^2)+I(slope^2)+I(btemp^2)+I(tmax^2)+I(speed^2)+I(color^2),data=trainc,family=gaussian)
summary(corals.GLM.cpue)

#reduced model
#fullmodel
corals.GLM.cpue<-glm((Coral_generic)^.25~lon+lat+depth+btemp+slope+color+I(lat^2)+I(btemp^2)+I(color^2),data=trainc,family=gaussian)
summary(corals.GLM.cpue)
#plot(corals.GLM.cpue,scale=0,residuals=TRUE)

#test the predictions on the training data
observed.train.GLM<-(trainc$Coral_generic)^.25
predicted.train.GLM<-predict(corals.GLM.cpue,trainc,type="response")
pred.train.GLM<-lm(observed.train.GLM~predicted.train.GLM)
summary(pred.train.GLM)
plot(observed.train.GLM,predicted.train.GLM)
abline(pred.train.GLM)

#test the predictions on the test data
observed.test.GLM<-(testc$Coral_generic)^.25
predicted.test.GLM<-predict(corals.GLM.cpue,testc,type="response")
pred.test.GLM<-lm(observed.test.GLM~predicted.test.GLM)
summary(pred.test.GLM)
plot(observed.test.GLM,predicted.test.GLM)
abline(pred.test.GLM)

temp_raster<-cut(corals.GLM.pa.raster,breaks=c(train.threshold,1))
corals.GLM.cpue.raster<-predict(raster.stack.GOA, corals.GLM.cpue,filename=paste(predictionrasters.path,"coralsGLMcpue",sep=""),fun=predict, na.rm=TRUE,overwrite=TRUE,progress="text",type="response",newdata.guaranteed=TRUE)

corals.GLM.cpue.raster<-mask(corals.GLM.cpue.raster,temp_raster)
writeRaster(corals.GLM.cpue.raster,filename=paste(predictionrasters.path,"coralsGLMcpue",sep=""),overwrite=TRUE)
plot(corals.GLM.cpue.raster, main="Coral GAM - PA, GAM for CPUE, Threshold = 0.05")
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)

#####OVERALL R2 GLM######
train.observed<-(train$Coral)^0.25
train.predicted<-predict(corals.GLM.pa,train,type="response")
train.predicted[train.predicted>=train.threshold]<-1
train.predicted[train.predicted<train.threshold]<-0
train.predicted.cpue<-predict(corals.GLM.cpue,train,type="response")
train.predicted.cpue<-train.predicted.cpue*train.predicted
# Chris: same fail as earlier
pred.train<-lm(train.predicted.cpue~train.observed)
summary(pred.train)
plot(train.observed,train.predicted.cpue)
abline(pred.train)
mse_coral<-sum(pred.train$residuals^2)/pred.train$df.residual
mse_coral

test.observed<-(test$Coral)^0.25
test.predicted<-predict(corals.GLM.pa,test,type="response")
test.predicted[test.predicted>=train.threshold]<-1
test.predicted[test.predicted<train.threshold]<-0
test.predicted.cpue<-predict(corals.GLM.cpue,test,type="response")
test.predicted.cpue<-test.predicted.cpue*test.predicted
pred.test<-lm(test.predicted.cpue~test.observed)
summary(pred.test)
plot(test.observed,test.predicted.cpue)
abline(pred.test)
mse_coral<-sum(pred.test$residuals^2)/pred.test$df.residual
mse_coral
```


### METHOD 4 - RANDOM FOREST METHOD
```{r, cache=TRUE, echo=FALSE}
corals.RF.pa<-randomForest(as.factor(corals_present)~lat+lon+depth+slope+btemp+speed+tmax+color,data=train, ntree=1000,importance=TRUE, na.action=na.omit)
print(corals.RF.pa)
round(importance(corals.RF.pa), 2)

train_predict_RF<-predict(corals.RF.pa,train,type="prob")[,2]
train.auc_data<-data.frame(cbind(seq(1,length(train_predict_RF),1),train[,"corals_present"],train_predict_RF))

#Calculate the AUC
auc(train.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
optimal.thresholds(train.auc_data,opt.methods=c(seq(1:9)))
train.threshold<-optimal.thresholds(train.auc_data,opt.methods=2)
train.threshold<-train.threshold[,2]
auc.roc.plot(train.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(train.auc_data,N.bins=10)
error.threshold.plot(train.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(train.auc_data,threshold=train.threshold)
pcc(cmx(train.auc_data,threshold=train.threshold))
sensitivity(cmx(train.auc_data,threshold=train.threshold))
specificity(cmx(train.auc_data,threshold=train.threshold))
predicted.prevalence(train.auc_data,threshold=train.threshold)
presence.absence.accuracy(train.auc_data,threshold=train.threshold)
presence.absence.hist(train.auc_data,color=c("green","red"),truncate.tallest=TRUE)
presence.absence.summary(train.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(train.auc_data[,2],train.auc_data[,3])


#Test on test data
test_predict_RF<-predict(corals.RF.pa,test,type="prob")[,2]
test.auc_data<-data.frame(cbind(seq(1,length(test_predict_RF),1),test[,"corals_present"],test_predict_RF))

#Calculate the AUC
auc(test.auc_data,na.rm=TRUE)

#Estimate the thresholds and calculate diagnostics
auc.roc.plot(test.auc_data,opt.methods=c(seq(1:9)))
calibration.plot(test.auc_data,N.bins=10)
error.threshold.plot(test.auc_data,opt.methods=c(seq(1:9)),opt.thresholds=TRUE)
cmx(test.auc_data,threshold=train.threshold)
pcc(cmx(test.auc_data,threshold=train.threshold))
sensitivity(cmx(test.auc_data,threshold=train.threshold))
specificity(cmx(test.auc_data,threshold=train.threshold))
predicted.prevalence(test.auc_data,threshold=train.threshold)
presence.absence.accuracy(test.auc_data,threshold=train.threshold)
presence.absence.hist(test.auc_data,color=c("green","red"),truncate.tallest=TRUE)  # random forest overfits the data
presence.absence.summary(test.auc_data,opt.methods=c(2,4,5),N.bins=10,truncate.tallest=TRUE)
cor.test(test.auc_data[,2],test.auc_data[,3])


#Predict to Raster
corals.RF.pa.raster<-predict(raster.stack.GOA, corals.RF.pa,type="prob",filename=paste(predictionrasters.path,"coralsRFpa",sep=""), na.rm=TRUE,overwrite=TRUE,progress="text",newdata.guaranteed=TRUE)
plot(1-corals.RF.pa.raster,main="Coral Boosted Regression Tree - PA")   # was the probability of the zero category
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)

#####RF CPUE MODEL
# no need to do the following
# Chris, is the correct transform?
 train$corals.fourth <- train$Coral^(1/4)
 test$corals.fourth <- test$Coral^(1/4)

corals.RF<-randomForest(corals.fourth~lat+lon+depth+slope+btemp+speed+tmax+color,data=train, ntree=1000,importance=TRUE, na.action=na.omit)
print(corals.RF)
round(importance(corals.RF), 2)

#Test on training data
train_predict_RF<-predict(corals.RF,train,type="response")
plot(train_predict_RF~train$corals.fourth)
RF.pred.train<-lm(train_predict_RF~train$corals.fourth)
summary(RF.pred.train)
abline(RF.pred.train)
mse_coral<-sum(RF.pred.train$residuals^2)/RF.pred.train$df.residual
mse_coral

#Test on test data
test_predict_RF<-predict(corals.RF,test,type="response")
plot(test_predict_RF~test$corals.fourth)
RF.pred.test<-lm(test_predict_RF~test$corals.fourth)
summary(RF.pred.test)
abline(RF.pred.test)
mse_coral<-sum(RF.pred.test$residuals^2)/RF.pred.test$df.residual
mse_coral

#Predict to Raster
corals.RF.cpue.raster<-predict(raster.stack.GOA, corals.RF,type="response",filename=paste(predictionrasters.path,"coralsRFcpue",sep=""), na.rm=TRUE,overwrite=TRUE,progress="text",newdata.guaranteed=TRUE)
#corals.RF.cpue.raster<-mask(corals.RF.cpue.raster,corals.habitat.mask,overwrite=TRUE)
#writeRaster(corals.RF.cpue.raster,filename=paste(predictionrasters.path,"coralsRFcpue",sep=""),overwrite=TRUE)
plot(corals.RF.cpue.raster,main="Coral Boosted Regression Tree - PA")
plot(akland,col=1,add=TRUE)
plot(canada_land,col=1,add=TRUE)

removeTmpFiles(h=.25)

```

